Numerical Investigation of Multi-Fluid Flows through Dual Saturated-Unsaturated Porous Media

### 2015-11-20 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
initial update and comments of the paper ...

\newcommand{\JGnote}[1]{\fbox{\parbox{\textwidth}{ \color{blue} JG Note $\Rightarrow$ #1}}}
\newcommand{\red}{\textcolor{red}}
\journal{Advances in Water Resources}

@@

%\medskip
Under certain \red{(spatial or geological, which is the best word?)} conditions the displacement of fluids in porous media is unstable and results in the formation of irregular flow patterns in the form of \textit{fingers}. This phenomenon, that describes these flow instabilities mentioned above is referred to as \textit{viscous fingering}. Even though the mechanism of the instabilities may vary (in immiscible fluids the most important factor is the surface tension while in miscible fluids, it is the diffusion \red{(thermal or mass diffusion ...be specific)}), however it is still a function of several variables such as, relative mobility, gravity, dispersion, permeability differences. Moreover, in most cases interfacial tension is directly linked with the mobility ratio (or viscosity ratio) between the displaced and the displacing fluid.

Both the evolution of visualisation techniques through the extensive use of parallel computing as well as the advances in modelling and simulation of flows in porous media may help to represent and address the importance of this phenomenon which can be found in petroleum, chemical and environmental related applications.

%\medskip
%\begin{comment}
Finite difference methods (FDM) have been extensively used in most industry-standard flow simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications (e.g., black oil model, passive tracer advection etc). However, FDM are often limited to relatively simple geometries (with low grid quality and orientation) representing idealised geological formations~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}. The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to overcome these deficiencies.

Among FEM-based formulations for porous media, the control volume finite element methods (CVFEM) ~\citep[CVFEM,][]{fung_1992,cordazzo_2004} has been largely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements~\citep{forsyth_1990, geiger_2004, hurtado_2007}. In this formulation, standard Galerkin-FEM method is used to discretise the pressure equation (FEM representation of pressure and velocity fields) whereas a control volume-based formulation is used to discretise all remaining scalar fields~\citep[e.g., saturation, density, species concentration etc, see][]{gomes_book_2012}.

Control volume finite element methods require high resolution meshes in regions where material properties vary abruptly, such as permeability and/or porosity contrasts at fracture-matrix interfaces. Since geometries are captured by finite elements (FE), constructed control volumes (CV) typically extend over the interfaces which may have different properties. Therefore, some average value of the permeability (defined in the FE space but projected into the CV space) is applied across the CVs at these interface. This often leads to excessive numerical dispersion especially in highly heterogeneous media. In order to overcome such artificial numerical dispersion, discontinuous hybrid finite element finite volume method (DFEFVM) was developed by \citet{nick_2011b, nick_2011a}. This novel discretisation scheme was designed to simulate flows through discrete fractured rocks in which CVs are divided along the interfaces of different materials.

\citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. Fluxes over CVs were calculated based upon material properties, whereas the saturation fields were volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.
%\end{comment}

%\medskip 
%\begin{center}
%\JGnote{WR: small summary of the paper -- 3 or 4 punchy sentences %followed by description of the remaining of the paper}
%\end{center}

In this paper, \textcolor{red}{the main objective is to investigate the impact of heterogeneous porous media in fingers formation using a FEM-based unstructured fixed mesh multi-fluid model with uniform permeability distribution under both saturated and unsaturated porous media}. The overlapping control volume finite element method (OCVFEM), formulation recently introduced by~\citet{jackson_2013}~\citep[see also][]{gomes_2013,pavlidis_2013} is used to investigate the development of multiphase flows through heterogeneous porous media.This finite element based method has a dual pressure-velocity representation in CV and FEM spaces  are embedded into a family of triangular/tetrahedral element pairs, P$_{m}$DG-P$_{n}$~\citep{cotter_2009b} and P$_{m}$DG-P$_{n}DG$~\citep{xie_2014}. The dual velocity and pressure fields are represented by $m^{\rm th}$- (discontinuous) and $({n})^{\rm th}$-order (continuous/discontinuous) polynomials, respectively.
Moreover, velocity field has dual mapping in FEM and CV spaces. This approach can achieve an exact representation of the Darcy force balance equations. 
In addition, the study focused in flow instabilities determined by the viscosity ratio. In the simulations presented in this paper the element pairs as described above and the overlapping control volume finite element method (CVFEM) model were used, in which the dual velocity-pressure fields are embedded in FEM space with simultaneous projection into CV space. Scalar fields (e.g., saturation, concentration, density, etc) are represented in the control volume space. High-order accurate downwind schemes on element boundaries on discontinuous scalar fields are flux-limited (based on NVD approach) to obtain bounded and compressive (capturing the interfaces) solutions. Section $2$ summarises the numerical formulation used in this work, followed by description of the flow instabilities  and a description of model validation. Finally, conclusions are drawn in the final section.

>>>>

\newcommand{\JGnote}[1]{\fbox{\parbox{\textwidth}{ \color{blue} JG Note $\Rightarrow$ #1}}}
\newcommand{\red}{\textcolor{red}}
\journal{Advances in Water Resources}
\newcommand{\frc}{\displaystyle\frac}
\newcommand{\PN}[2][error]{P$_{#1}$DG-P$_{#2}$}
\newcommand{\PNDG}[2][error]{P$_{#1}$DG-P$_{#2}$DG}

@@

%\medskip
Under specfic spatial conditions the displacement of fluids in porous media is unstable {\bf(why?)} and results in the formation of irregular flow patterns in the form of fingers. This phenomenon, viscous fingering, describes flow instabilities {\bf (and ...??)}. Even though the mechanism of instabilities may vary (in immiscible fluids the most important factor is the surface tension while in miscible fluids, it is the diffusion \red{(thermal or mass diffusion ...be specific)}), however it is still a function of several variables such as, relative mobility, gravity, dispersion, permeability differences. Moreover, in most cases interfacial tension is directly linked with the mobility ratio (or viscosity ratio) between the displaced and the displacing fluid.
\medskip
%Both the evolution of visualisation techniques through the extensive use of parallel computing as well as the advances in modelling and simulation of flows in porous media may help to represent and address the importance of this phenomenon which can be found in petroleum, chemical and environmental related applications. {\bf (does the paper address any aspect of parallel computation or computational/data visualization?)}

In this paper, the overlapping control volume finite element method (CVFEM) formulation, recently introduced by~\citet{jackson_2015}~\citep[see also][]{radunz_2014}, is used to investigate the development of multiphase flows through heterogeneous porous media. In this formulation, the dual pressure-velocity representation in CV and FEM spaces are embedded into a family of triangular/tetrahedral element-pairs, \PN[n]{m} and \PNDG[n]{m} \citep[][]{cotter_2009b,xie_2014} -- basis function with $n^{\rm th}$- (discontinuous) and $m^{\rm th}$-order (continuous/discontinuous) polynomials for velocity and pressure fileds, respectively.  Scalar fields (e.g., saturation, concentration, density, etc) are represented in the CV space. High-order accurate downwind schemes on element boundaries on discontinuous scalar fields are flux-limited (based on NVD approach) to obtain bounded and compressive (capturing the interfaces) solutions. This formulation can achieve an exact representation of the Darcy force balance equations.

\medskip

Here, the main objective is to investigate the impact of heterogeneous porous media in fingers formation using a FEM-based unstructured fixed mesh multi-fluid model with uniform permeability distribution under both saturated and unsaturated porous media. {\bf (try to develop it further; be more specific. Write this objective sentence/paragraph based on the paper's conclusions!)}. This study focuses on flow instabilities determined {\bf(determined or controlled?)} by the viscosity ratio.  Section $2$ summarises the numerical formulation used in this work, followed by description of the flow instabilities  and a description of model validation. Finally, conclusions are drawn in the final section. {\bf (Needs better description of the remaining of teh paper.)}
